\begin{singlespace}

Demmocritus was known as the laughing philosopher while Heraclitus, in contrast, was known as both the weeping philosopher  and \emph{the obscure}. There are many people I'd like to thank for making my experience and the end result of my time in graduate school closer, in many senses, to the former rather than the latter.  At this point, I can't help but take the perspective of Heraclitus with the attitude of Demmocritus.

First, I'd like to thank my advisor Robin Clark for his open-mindedness, encouragement, and patience. The best of this dissertation has come from hours of discussion that have never failed to broaden my view of linguistics and guidance that continues to strike me as far-sighted. I'd also like to thank Mark Liberman and Florian Schwarz for their suggestions and perspectives that greatly improved this work. More broadly, I'd like to thank the faculty who have influenced my thinking through various conversations, including Tony Kroch, Bill Labov, Don Ringe, Gillian Sankoff, and Charles Yang.

Second, I want to acknowledge several individuals without whom this dissertation would not have been possible. I'd like to thank Jon Stevens for planting the broader idea of this project in my head at the outset of graduate school. I've benefitted immeasurably from our discussions over the years and hope this is a reasonable first approximation to some of those ideas. I'd also like to thank Aaron Ecay for sharing code, data, and an office. I've benefitted immensely from his generosity of patience and expertise. I'd also like to thank Josh Plotkin and Mitchell Johnson for sharing code and helping me gain a better understanding of population dynamics in finite populations. I'd also like to thank Joe Fruehwald, Kyle Gorman, Soohyun Kwon, Constantine Lignos,  Hilary Prichard, Kobey Shwayder, and Meredith Tamminga for advice and encouragement at important junctures in graduate school. 

Third, I'd like to thank  Amy Forsyth, Laurel Sweeney, and Jessica Marcus who have made navigating the administrative side of things feel effortless. Without exaggeration, their support made getting through the non-administrative side of graduate school possible.

Finally, and most importantly, I'd like to thank my family. Mom, Dad, Opa, David, Matthew, and Sarah, thank you. Your love and encouragement have made me who I am. Rebecca, thank you. The best of who I am stems from your love and patience.
 

\end{singlespace}
